---
title: "CHRIS MS-based plasma proteome: general overview and sex, age, BMI associated proteins"
author: "Nikola Dordevic, Johannes Rainer"
graphics: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: hide
bibliography: references.bib
---

**Modified**: `r file.info("sex_age_bmi_associated_proteins.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r biocstyle, echo = FALSE, results = "hide", message = FALSE}
#' rmarkdown format settings
library(BiocStyle)
library(knitr)
BiocStyle::markdown()
```

```{r settings, echo = FALSE}
#' General settings
filename <- "sex_age_bmi_associated_proteins"
#' Path to save images to
IMAGE_PATH <- paste0("images/", filename, "/")
if (file.exists(IMAGE_PATH))
    unlink(IMAGE_PATH, recursive = TRUE)
dir.create(IMAGE_PATH, recursive = TRUE)
#' Path to store RData files
RDATA_PATH <- paste0("data/RData/", filename, "/")
dir.create(RDATA_PATH, recursive = TRUE, showWarnings = FALSE)
#' wheter interactive plots should be used. Should be set to FALSE if report
#' is to be shared with external researchers.
PLOTLY <- FALSE
ggplotly_or_not <- function(x, interactive = TRUE) {
    if (interactive) ggplotly(x)
    else print(x)
}
opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8,
               fig.height = 7, fig.path = IMAGE_PATH)
chris_path <- "/home/jo/data/CHRIS/chris_baseline/"
```

# Introduction

This document provides a general description of the CHRIS ScanningSWATH mass
spectrometry (MS)-based plasma proteome data set and describes the analysis to
identify plasma proteins that are associated to age, sex and body mass index
(BMI). The analysis is performed similarly to the analysis for the
identification of age, sex and BMI-associated metabolites
[@verri_hernandes_age_2022].

# Data import

Below we load the data and all required libraries. The data is loaded from the
CHRIS TDFF modules.

```{r data-libraries}
library(chrisUtils)
library(chrisData)
library(RColorBrewer)
library(pander)
library(pheatmap)
library(DT)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(plotly)
options(rgl.useNULL = TRUE)
library(rgl)
library(readxl)
library(writexl)
library(tidyfr)
mdl <- data_module("chris_scanningswath_proteins", "1.0.0.1", chris_path)
prot_data <- data(mdl)
prot_ann <- labels(mdl)
stopifnot(all(colnames(prot_data) == rownames(prot_ann)))
colnames(prot_data) <- prot_ann$description
rownames(prot_ann) <- prot_ann$description
## define the columns with protein concentrations
proteins <- rownames(prot_ann)[prot_ann$Genes != ""]
```

Before proceeding with the data analysis we remove the POOL samples. In addition
we restrict to samples from the CHRIS study.

```{r remove-POOLs-NAFLDs, message = FALSE, warning = FALSE}
prot_data <- prot_data[grep("^001", rownames(prot_data)), ]
```

Next we define the various variables which will be included in the analysis.
First we define and add the BMI data, both as a numeric value, but also using
the BMI categories. For the BMI categories we are using *2* (18.5 - <25) as
baseline since that is supposed to be the *normal*.

```{r define-variables-bmi, message = FALSE, warning = FALSE}
clin <- chrisData("clinical")
rownames(clin) <- clin$AID
## Adding this information to the colData
prot_data$BMI <- clin[rownames(prot_data), "x0an03q"]
prot_data$BMIcat <- factor(as.integer(clin[rownames(prot_data), "x0an03b"]),
                           levels = c("2", "1", "3", "4"))
```

We also define the season in which samples were collected from the study
participants and add the age and sex information for each study participant.

```{r define-variables-general-info, message = FALSE}
## add here phenotype data from chrisData and recalculate season from exam day
gen_info <- chrisData("general_information", release = "baseline")
rownames(gen_info) <- gen_info$AID
prot_data$Season <- as.factor(season(gen_info[rownames(prot_data), "x0_examd"]))
prot_data$Sex <- gen_info[rownames(prot_data), "x0_sex"]
prot_data$Age <- gen_info[rownames(prot_data), "x0_age"]
```

```{r define-variables-pregnant, message = FALSE}
interview_info <- chrisData("interview", release = "baseline")
rownames(interview_info) <- interview_info$AID
## Add pregnancy information
prot_data$Pregnant <- interview_info[rownames(prot_data), "x0wo05"]
```

A categorical variable for fasting status is defined based self-reported data
from the participants' interview.

```{r define-variables-fasting, message = FALSE}
## add number of people reporting fasting status (`x0bc12` from `chrisData`
## `"labdata"`)
labdata_info <- chrisData("labdata", release = "baseline")
rownames(labdata_info) <- labdata_info$AID
prot_data$Fasting <- labdata_info[rownames(prot_data), "x0bc12"]
```

At last we load also the medication data which might be used later to test for
potential confounding.

```{r}
medi <- chrisData("drugs_1")
library(atc)
```


# General data overview

In resulting dataset, there are `r nrow(prot_data)` samples in total and
general overview of the trait is summarized below:

```{r, message=FALSE}
Overview <- summary(
    prot_data[, c("Sex", "Fasting", "Age", "Season", "BMIcat",
                  "Pregnant", "Fasting")])
print(Overview)
```

We further remove all individuals with missing values for any of the required
variables (age, sex, fasting and BMI) as well as pregnant women.

```{r}
nas <- is.na(prot_data[, c("Sex", "Fasting", "Age", "BMIcat")])
prot_data <- prot_data[rowSums(nas) == 0, ]
## remove pregnant women
rem <- c(which(prot_data$Pregnant == "Yes"),
         which(prot_data$Pregnant == "Don't know, possible"))
prot_data <- prot_data[-rem, ]
```

Summary for the final data set.

```{r}
prot_data[, c("Sex", "Fasting", "Age", "Season", "BMIcat",
              "Pregnant", "Fasting")] %>% summary()
```

This reduces the data set to `r nrow(prot_data)` individuals.

In next figure, scores and loadings from the PCA are presented considering all
available `r length(proteins)` proteins.
Samples are coloured according to their sex. Prior the PCA, data is normalized
to 0 mean and standard deviation of 1 (autoscaling).
Plot shows the scores and loadings of all subjects in the space of the first two
PCs, with 15% of total variances explained.

```{r PCA_1_figure, fig.path = IMAGE_PATH, fig.width=12, fig.height=7}
## include age in PCA
scaledData <- scale(prot_data[, proteins], scale = TRUE)
scPCA <- svd(scaledData)
scPCA.scores <- scPCA$u %*% diag(scPCA$d)    ## scores
scPCA.loadings <- scPCA$v                    ## loadings
rownames(scPCA.loadings) <- proteins
scPCA.variances <- round(100*((scPCA$d^2) / sum(scPCA$d^2)), 1)[1:10]
Sex <- prot_data$Sex
gg_1 <- autoplot(prcomp(scaledData), data = prot_data,
                 loadings = TRUE, loadings.colour = 'gray',
                 loadings.label = TRUE, loadings.label.size = 3,
                 loadings.label.colour ="darkblue", colour = NA) +
    geom_point(aes(colour = Sex)) +
    theme_bw()
ggplotly_or_not(gg_1, PLOTLY)
```

```{r}
## FIGURE 1
## Transparency for protein names depending on arrow length?
png("images/manuscript/Figure_2.png", width = 10, height = 10, units = "cm",
    res = 600, pointsize = 4)

pc <- prcomp(scaledData)
library(CompMetaboTools)
cols <- rep("#E41A1C", nrow(prot_data))
cols[prot_data$Sex == "Male"] <- "#377EB8"

layout(matrix(1:2), heights = c(1, 0.25))
par(mar = c(4.5, 4.3, 0.5, 0.5), cex.lab = 1.5, bty = "n")

plot_pca(pc, pch = NA, col = NA, bg = paste0(cols, 60), cex = 1.5,
         xlim = c(-16, 9))

unsigned.range <- function(x)
        c(-abs(min(x, na.rm = TRUE)), abs(max(x, na.rm = TRUE)))
y <- pc$rotation[, 1:2]
scl <- max(
    unsigned.range(y[, 1]) / unsigned.range(pc$x[, 1]),
    unsigned.range(y[, 2]) / unsigned.range(pc$x[, 2]))
y <- y / scl
arrows(x0 = 0, x1 = y[, 1],
       y0 = 0, y1 = y[, 2],
       lwd = 0.75, angle = 30, length = 0.05,
       col = "#00000050")
text(y[, 1] * 1.1, y[, 2] * 1.1, labels = rownames(y),
     col = "#00000080", cex = 1.4)

points(pc$x[, 1], pc$x[, 2], pch = 21, col = NA, bg = paste0(cols, 60),
       cex = 1.5)
grid()

## Age distribution along x.
brks <- seq(-16, to = 9, by = 1)
f <- cut(pc$x[, 1], breaks = brks)
ages <- split(prot_data$Age, f = f)
par(mar = c(0.5, 4.3, 0.5, 0.5))
boxplot(ages, at = brks[-1] - 0.5, xlim = c(-16, 9), xaxt = "n",
        ylab = "Age", lwd = 0.75)
grid()

dev.off()
```


There is an interesting separation on PC1 with a subgroup of mostly female
participants which seems also to be related to Age (not shown).

If we rotate the figure, it would be clear that separation between males and
females is more clear if we consider PC1 and PC3:

```{r PCA_2_figure, fig.path = IMAGE_PATH, fig.width=12, fig.height=7}
## include age in PCA
gg_2 <- autoplot(prcomp(scaledData), x = 1, y = 3,
                 data = prot_data, loadings = TRUE, loadings.colour = 'gray',
                 loadings.label = TRUE, loadings.label.size = 3,
                 loadings.label.colour = "darkblue") +
    geom_point(aes(colour = Sex), size = 1) +
    theme_bw ()
ggplotly_or_not(gg_2, PLOTLY)
```

From the scores, there is a clear effect of sex in direction of PC1. From the
loadings we can observe relationships between proteins. Every arrow corresponds
to the one parameter (protein) in data set. Longest arrows in the direction of
particular Principal Component have largest impact and importance for that PC.
The same direction of groups of arrows and their lengths indicates correlation
of corresponding parameters in the data set and opposite: opposite directed
arrows indicate negative correlation between corresponding parameters. For the
cases when angles between arrows are around 90°, indicates that there is no
correlation between corresponding parameters.
Note that with the large numbers of parameters in PCA, some loadings can be
placed close to each other even if there is no strong correlation. Thus,
attention is necessary.

The protein P01019 (AGT, Angiotensinogen) seems to be one driver for the
separation of a set of (younger) female participants from the bulk of male and
female participants. Since this protein was shown in the exploratory analysis to
be eventually dependent on medication with sexual hormones we highlight
participants with that treatment in the PCA below.

We next define the categorical variable for *hormonal contraceptives* using the
ATC level 3 medication information from CHRIS. In particular, we define the
trait using the ATC3 categories *HORMONAL CONTRACEPTIVES FOR SYSTEMIC USE* and
*ANTIANDROGENS*. This definition is similar to the one in Ramsey et al
[@ramsey_variation_2016] that considered women taking oral contraceptives.

```{r, warning = FALSE}
## approach 1
med <- chrisData("drugs_1")
med <- med[med$AID %in% rownames(prot_data), ]
med <- med[!is.na(med$x0dd03), ]
med$atc_2 <- toAtcLevel(med$x0dd03, level = 2)
med$atc_3 <- toAtcLevel(med$x0dd03, level = 3)
med$atc_4 <- toAtcLevel(med$x0dd03, level = 4)
atc_df <- as.data.frame(atc)
med$atc_2_name <- atc_df[match(med$atc_2, atc_df$key), "name"]
med$atc_3_name <- atc_df[match(med$atc_3, atc_df$key), "name"]
med$atc_4_name <- atc_df[match(med$atc_4, atc_df$key), "name"]
## Subset to selected sex hormone treatment.
med_sex_horm <- med[med$atc_3_name %in%
                    c("HORMONAL CONTRACEPTIVES FOR SYSTEMIC USE",
                      "ANTIANDROGENS"), ]
prot_data$Sex_horm <- "No"
prot_data$Sex_horm[match(med_sex_horm$AID, rownames(prot_data))] <- "Yes"
prot_data$Sex_horm <- factor(prot_data$Sex_horm,
                             levels = c("No", "Yes"))
summary(prot_data$Sex_horm)
```


```{r PCA_sex_hormone, warning = FALSE, fig.width = 12, fig.height = 7}
gg_2 <- autoplot(prcomp(scaledData), x = 1, y = 2,
                 data = prot_data, loadings = TRUE, loadings.colour = 'gray',
                 loadings.label = TRUE, loadings.label.size = 3,
                 loadings.label.colour = "darkblue", colour = NA) +
    geom_point(aes(colour = Sex_horm), size = 1) +
    theme_bw ()
ggplotly_or_not(gg_2, PLOTLY)
```

```{r, echo = FALSE}
## Supplementary Figure PCA
## Same plot but with coloring representing HCU.
png("images/manuscript/supplement/Figure_S149.png", width = 8.2, height = 6.4, units = "cm",
    res = 600, pointsize = 4)

cols <- rep("#999999", nrow(prot_data))
cols[prot_data$Sex_horm == "Yes"] <- "#FF7F00"

par(mar = c(4.5, 4.3, 0.5, 0.5), cex.lab = 1.5, bty = "n")

plot_pca(pc, pch = NA, col = NA, bg = paste0(cols, 60), cex = 1.5,
         xlim = c(-16, 9))

unsigned.range <- function(x)
        c(-abs(min(x, na.rm = TRUE)), abs(max(x, na.rm = TRUE)))
y <- pc$rotation[, 1:2]
scl <- max(
    unsigned.range(y[, 1]) / unsigned.range(pc$x[, 1]),
    unsigned.range(y[, 2]) / unsigned.range(pc$x[, 2]))
y <- y / scl
arrows(x0 = 0, x1 = y[, 1],
       y0 = 0, y1 = y[, 2],
       lwd = 0.75, angle = 30, length = 0.05,
       col = "#00000050")
text(y[, 1] * 1.1, y[, 2] * 1.1, labels = rownames(y),
     col = "#00000080", cex = 1.4)

points(pc$x[, 1], pc$x[, 2], pch = 21, col = NA, bg = paste0(cols, 60),
       cex = 1.5)
grid()
dev.off()

## Supplementary Figure P01019
f <- as.character(prot_data$Sex)
f[prot_data$Sex_horm == "Yes"] <- "HCU"
f <- factor(f, levels = c("HCU", "Female", "Male"))

png("images/manuscript/supplement/Figure_S150.png", width = 6.4, height = 6.4, units = "cm",
    res = 600, pointsize = 4)
library(vioplot)
par(mar = c(4.3, 4.5, 2, 0.5), cex.axis = 1.5, cex.lab = 1.5, bty = "n",
    cex.main = 2.5)
vioplot(split(prot_data$P01019, f), ylab = expression(log[2]~abundance),
        main = "P01019 (AGT)")
grid(nx = NA, ny = NULL)
dev.off()
```

Most of the participants separating from the main group take indeed systemic
contraceptives. For participants in that group that are not assigned to that
trait is is likely that they did not bring the medication (contraceptives) with
them to the study center and thus no ATC code was scanned.

In next box-blot, age distribution per sex is visualized. Points are colored
according the "Sex_horm" variable:

```{r, BP_age_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
prot_plot <- function(data, aes, protein) {
    ggplot(data, aes) +
        geom_jitter(position = position_jitter(0.24),
                    aes(colour = Sex_horm, alpha = 0.5)) +
        geom_violin(trim = FALSE, aes(alpha = 0.5)) +
        stat_summary(fun = median, geom = "point", shape = 23, size = 2) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8),
              axis.title.y = element_blank()) + labs(title = protein) +
        theme_bw ()
}
age_1 <- prot_plot(prot_data, aes(x = Sex, y = Age),
                   protein = "Sex Vs. Age")
ggplotly_or_not(age_1, PLOTLY)
```

Also, age distribution per Sex_horm is visualized:

```{r, BP_age_2_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
age_2 <- prot_plot(prot_data, aes(x = Sex_horm, y = Age),
                   protein = "Sex_horm Vs. Age")
ggplotly_or_not(age_2, PLOTLY)
```

Clearly there is an enrichment of young women among participants treated with
sexual hormones.


# Linear Regression analyses

To identify proteins related with sex, age, fasting status and BMI, linear
regression models are fitted for all proteins separately (as a response) and
sex, age, fasting status and (categorical) BMI as a covariates (same as in
[@verri_hernandes_age_2022]). In addition, and based on the observations from
the previous section, a categorical variable representing use of hormonal
contraceptives is included to adjust for the influence of this factor. Model
coefficients and p-values are provided. Raw p-values are adjusted for multiple
hypothesis testing with the method from Bonferroni.

```{r, message=FALSE}
## make a function for LM
FunForLinReg <- function(AnalyteName, Data) {
    Analyte <- Data[, AnalyteName]
    LMReg <- lm(Analyte ~ Sex + Age + Fasting + BMIcat + Sex_horm,
                data = Data)
    res <- coef(summary(LMReg))[-1, ]
    p.Values <- data.frame(c(res[, 1],
                             res[, 4]))
    names(p.Values) <- AnalyteName
    rownames(p.Values) <- c(paste0("coef_", rownames(res)),
                            paste0("p-value_", rownames(res)))
    return(p.Values)
}
## do it in one run
LMRegtests <- do.call(
    cbind, lapply(proteins, function(x) FunForLinReg(x, prot_data)))
Test_output <- t(LMRegtests)
## Adjusting for multiple hypothesis testing
adjp <- apply(Test_output[, grep("p-value", colnames(Test_output))],
              MARGIN = 2, p.adjust, method = "bonferroni")
colnames(adjp) <- sub("value", "adj", colnames(adjp))
Test_output <- cbind(Test_output, adjp)
```

To obtain comparable coefficients, prior to linear modelling all the variables
are standardized to zero mean and unit variance (note that this does not
influence the linear models since they are affine equivariant).
In autoscaled data, difference of 1 corresponds to standard deviation of 1.
This parameter will be called "effect size" and will be use to evaluate group
differences.

```{r, message = FALSE}
## perform data normalization (autoscaling)
prot_data_AS <- prot_data
prot_data_AS[, proteins] <- scale(prot_data_AS[, proteins], scale = TRUE)
## do it in one run
LMRegtests_AS <- do.call(cbind.data.frame,
                         lapply(proteins,
                                function(x) FunForLinReg(x, prot_data_AS)))
Test_output_AS <- t(LMRegtests_AS)
EffectSize <- Test_output_AS[, grep("coef", colnames(Test_output_AS))]
colnames(EffectSize) <- sub("coef", "effect_size", colnames(EffectSize))
```

We next define criteria to call proteins *significant*. Due the size of the data
set proteins would be called significant, even if their difference in
concentrations is only very small. Hence we evaluate whether we could combine
the criteria for *statistical significance* (the p-value) with another criteria
that would allow us to define the proteins that are the *strongest* associated
with the contrast of interest.

```{r significant-effect-size}
comps <- colnames(Test_output)[grep("coef", colnames(Test_output))]
sign_p <- Test_output[, grep("p-adj", colnames(Test_output))] < 0.05
colnames(sign_p) <- sub("p-adj", "significant", colnames(sign_p))
## Define significant analytes based on effect size.
cut_effect_size <- 0.5
sign_es <- sign_p &
    abs(EffectSize) > cut_effect_size
sign_es[, "significant_Age"] <- sign_p[, "significant_Age"]
```

We require in addition to the *statistical significance* that the difference in
concentrations for a protein is larger than its *technical variance*. Technical
variance of an protein is estimated by the coefficient of variation of the
protein in QC CHRIS Pool samples.

Below we add thus evaluate for each protein and comparison whether it fulfills
that criteria.

```{r define-significant}
#' Calculate the (absolute) difference in abundance expressed as a percentage
#' based on a (log2) coefficient.
diff_percentage <- function(x) {
    (2^abs(x) - 1) * 100
}
comps <- colnames(Test_output)[grep("coef", colnames(Test_output))]
diff_perc <- apply(Test_output[, comps],
                   MARGIN = 2, diff_percentage)
colnames(diff_perc) <- sub("coef", "diff_perc", colnames(diff_perc))
#' define which metabolites are considered significant: require p-value < 0.05
#' and difference in concentration > CV_QC (expressed in %)
sign_cv <- lapply(comps, function(z) {
    z <- sub("coef_", "", z)
    sign_p[, paste0("significant_", z)] &
        diff_perc[, paste0("diff_perc_", z)] > 100 *
        prot_ann[proteins, "cv_qc_chris"]
})
sign_cv <- do.call(cbind, sign_cv)
colnames(sign_cv) <- sub("coef", "significant", comps)
#' Add the significant for numeric variable
sign_cv[, "significant_Age"] <- sign_p[, "significant_Age"]
```

The number of proteins called significant or with an adjusted p-value smaller
than 0.05 for each of the extracted coefficients are shown below.

```{r table-sig-mets, results = "asis"}
sig_tab <- data.frame(
    `p_adj < 0.05` = colSums(sign_p),
    cv = colSums(sign_cv),
    check.names = FALSE)
pandoc.table(
    data.frame(no_sign = sig_tab, check.names = FALSE),
    style = "rmarkdown",
    caption = paste0("Number of significant proteins for each ",
                     "comparison and definition of significance.",
                     "cv: difference in concentration > * CV in QC samples."))
```

```{r export-results, warning = FALSE}
## Loading also the information whether an analyte was already identified
## before to be significant.
results <- data.frame(
    prot_ann[proteins, c("description", "long_description", "PeptideNN",
                         "PeptideNames", "Protein.Ids", "Protein.Names",
                         "Genes", "Modified.Sequence", "Stripped.Sequence",
                         "Precursor.Id", "cv_qc_chris")],
    Test_output,
    EffectSize,
    diff_perc,
    sign_cv)
## Add average concentrations for most important groups.
results$Avg_Male <- colMeans(
    prot_data[which(prot_data$Sex == "Male"), proteins])
results$Avg_Female <- colMeans(
    prot_data[which(prot_data$Sex == "Female"), proteins])
results$Avg_BMI1 <- colMeans(
    prot_data[which(prot_data$BMIcat == "1"), proteins])
results$Avg_BMI2 <- colMeans(
    prot_data[which(prot_data$BMIcat == "2"), proteins])
results$Avg_BMI3 <- colMeans(
    prot_data[which(prot_data$BMIcat == "3"), proteins])
results$Avg_BMI4 <- colMeans(
    prot_data[which(prot_data$BMIcat == "4"), proteins])
dr <- paste0("data/xlsx/", filename)
dir.create(dr, showWarnings = FALSE, recursive = TRUE)

colnames(results) <- sub("^description$", "Uniprot", colnames(results))
colnames(results) <- sub("^long_description$", "Description", colnames(results))
write_xlsx(
    results, path = paste0(dr, "/results_sex_age_bmi_sex_hormone_adjusted.xlsx"))
```


## Sex-associated plasma proteins

To identify sex related proteins, results from linear models fitted previously
are used.

```{r FIG-volcano-sex, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot for the sex-comparison. Red colored points are those called *significant* by considering the adjusted p-value and CV.", results = "hide", echo = FALSE}
plot_volcano <- function(x, comparison = character(), ...) {
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    plot(X, -log10(Y), pch = 21,
         col = "#000000ce", bg = "#00000060",
         xlab = "", ylab = "", xaxt = "n",
         ...)
    xat <- axTicks(1, usr = par("usr")[1:2])
    labs <- gsub("-", "\U2212", print.default(xat))
    axis(1, at = xat, labels = labs)
    mtext(side = 1, line = 2.2, text = "coefficient", cex = par("cex.axis"))
    mtext(side = 2, line = 2.3, text = expression(-log[10](p[adj])),
          cex = par("cex.axis"))
}
significant_points <- function(x, comparison = character(),
                               diff_perc = 2, col = "#E41A1CCE",
                               bg = "#E41A1C40", ...) {
    sig <- x[, paste0("significant_", comparison)]
    X <- x[, paste0("coef_", comparison)]
    Y <- x[, paste0("p.adj_", comparison)]
    Y[Y == 0] <- min(Y > 0) / 10
    if (any(sig))
        points(X[sig], -log10(Y[sig]), col = col, bg = bg, pch = 21)
}
plot_volcano(results, "SexFemale", main = "Female vs Male")
grid()
abline(h = -log10(0.05), lty = 2, col = "black")
significant_points(results, comparison = "SexFemale")
```

The (static) table with the significant proteins is shown below:

```{r table-sig-sex, results = "asis", echo = FALSE}
significant_table <- function(x, comparison = character(),
                              columns = c("Genes", "Description",
                                          "cv_qc_chris")) {
 sigs <- x[, paste0("significant_", comparison)]
 columns <- c(columns, paste0(c("coef_", "p.adj_", "effect_size_"), comparison))
 tmp <- x[sigs, columns]
 tmp <- tmp[order(tmp[, paste0("p.adj_", comparison)]), ]
 colnames(tmp) <- sub(paste0("_", comparison), "", colnames(tmp))
 tmp
}
sig_tab <- significant_table(results, "SexFemale")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_SexFemale.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for Sex (Female vs Male)."))
```

Top 3 proteins according to tests p-values are visualized in boxplots in next
figures. To evaluate influence of sexual hormone medication data points are
colored accordingly.

```{r, BP_Sex_1_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6, warning = FALSE}
s_1 <- prot_plot(prot_data, aes(x = Sex, y = P02753),
                   protein = "P02753")
ggplotly_or_not(s_1, PLOTLY)
```

Male study participants have an about 25% higher concentrations of RBP4 in
serum. Retinol-binding protein 4 belongs to the lipocaling family and is the
specific carrier for retinol (vitamin A alcohol) in the blood. This protein has
a significant role in the visual cycle as it is the primary transporter of
retinol to the retinal pigment epithelium. Serum retinol concentrations as well
as RBP4 were reported to be lower in females compared to male
[PMID:35108705](https://pubmed.ncbi.nlm.nih.gov/35108705/) (small, potentially
biased, cohort). This is thus not really a novel finding.

```{r, BP_Sex_2_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
s_2 <- prot_plot(prot_data, aes(x = Sex, y = P01871),
                 protein = "P01871")
ggplotly_or_not(s_2, PLOTLY)
```

Female study participants have an abuot 25% lower concentrations of IGHM in
serum. Immonuglobulins are the antigen recognition molecules of B-cells.

```{r, BP_Sex_3_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
s_3 <- prot_plot(prot_data, aes(x = Sex, y = P01861),
                 protein = "P01861")
ggplotly_or_not(s_3, PLOTLY)
```

In contrast to IGHM, IGHG4 has about 30% higher abundance in serum of male study
participants than of female study participants. Immunoglobin heavy constant
gamma 4 is predicted to enable antigen binding activity and immunoglobulin
receptor binding activity.


## Age-associated plasma proteins


To identify age-related proteins, results from linear models fitted previously
are used.

```{r FIG-volcano-age, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot showing age-related associations. Red colored points are those with an adjusted p-value smaller than 0.05.", echo = FALSE, results = "hide"}
plot_volcano(results, "Age", main = "Age")
abline(h = -log10(0.05), lty = 2)
grid()
significant_points(cbind(results, sign_cv), comparison = "Age")
```

```{r table-sig-age, echo = FALSE, results = "asis"}
sig_tab <- significant_table(results, "Age")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_Age.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for Age. "))
```

Top 3 proteins according to tests p-values are visualized in scater plots with
age and colored according sex.
On the top, pearson correlation coefficients are present with their p-values.


```{r, SP_Age_1_figure, fig.path = IMAGE_PATH, fig.width = 6, fig.height = 5}
scatter_plot <- function(data, aes, protein) {
    ggplot(data, aes) +
    geom_jitter(aes(colour = Sex, alpha = 0.8),
    position=position_jitter(0.05)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    panel.background = element_blank()) +
    ggpubr::stat_cor(method = "pearson", label.x.npc = "middle",
    label.y.npc = "top", col="black") + labs(title = protein) +
    theme_bw()
    }
a_1 <- scatter_plot(prot_data, aes(x = Age, y = `P35858;P35858-2`),
                    protein  = "P35858;P35858-2")
ggplotly_or_not(a_1, PLOTLY)
```

```{r, SP_Age_2_figure, fig.path = IMAGE_PATH, fig.width = 6, fig.height = 5}
a_2 <- scatter_plot(prot_data, aes(x = Age, y = `P04004`), protein  = "P04004")
ggplotly_or_not(a_2, PLOTLY)
```

```{r, SP_Age_3_figure, fig.path = IMAGE_PATH, fig.width = 6, fig.height = 5}
a_3 <- scatter_plot(prot_data, aes(x = Age, y = P02787), protein  = "P02787")
ggplotly_or_not(a_3, PLOTLY)
```

**IGFALS** (insulin like growth factor binding protein acid labile subunit) is a
serum protein that binds insulin-like growth factors, increasing their half-life
and their vascular localization. Downregulation of IGFALS mRNA with age was
described on a two-group liver transcriptome analysis
[PMID:34959291](https://pubmed.ncbi.nlm.nih.gov/34959291/). Other reports
mentioning an age-dependency (or dependency on age-related traits) exist: [PMID:27329260](https://pubmed.ncbi.nlm.nih.gov/27329260/).

**VTN** (vitronectin) acts in part as an adhesive glycoprotein. This protein
also inhibits the membrane-damaging effect of the terminal cytolytic complement
pathway and binds to several serpin serine protease inhibitor. It is involved in
a variety of other biological processes such as the regulation of the
coagulation pathway, wound healing, and tissue remodeling. It is also a lipid
binding protein that forms a principal component of the high density
lipoprotein.

**APOB** (apolipoprotein B) is the main apolipoprotein of chylomicrons and low
density lipoprotein (LDL).

**APOL1** (apolipoprotein L1) is the major apoprotein of HDL.

Decreasing VTN and APOL1 with age might indicate decreasing concentration of HDL
with age, while LDL increases.


### Identification of BMI related proteins

To identify BMI related proteins, results from linear models fitted previously
are used. The volcano plot below shows the results for the 3 comparisons (i.e.
BMI category 1, 3 or 4 against the baseline BMI category 2).

```{r FIG-volcano-bmi, fig.path = IMAGE_PATH, fig.width = 15, fig.height = 5, fig.cap = "Volcano plot for BMI-related proteins. Red colored points are those with an adjusted p-value smaller than 0.05 and a difference in concentrations which is larger than 2 times the coefficient of variation for that analyte calculated on QC samples.", results = "hide", echo = FALSE}
par(mfrow = c(1, 3), mar = c(4, 4, 1, 0.5))
xlim <- range(results[, c("coef_BMIcat1", "coef_BMIcat3", "coef_BMIcat4")])
ylim <- range(
    -log10(results[, c("p.adj_BMIcat1", "p.adj_BMIcat3", "p.adj_BMIcat4")]))
plot_volcano(results, "BMIcat1", main = "BMI 1 against 2",
             xlim = xlim, ylim = ylim)
abline(h = -log10(0.05), lty = 2)
grid()
significant_points(results, sign_es, comparison = "BMIcat1")
plot_volcano(results, "BMIcat3", main = "BMI 3 against 2",
             xlim = xlim, ylim = ylim)
abline(h = -log10(0.05), lty = 2)
grid()
significant_points(results, sign_es, comparison = "BMIcat3")
plot_volcano(results, "BMIcat4", main = "BMI 4 against 2",
             xlim = xlim, ylim = ylim)
abline(h = -log10(0.05), lty = 2)
grid()
significant_points(results, sign_es, comparison = "BMIcat4")
```

As expected, the highest number of significant proteins was detected for the
comparison of BMI category 4 against baseline 2. Also, the extent and
significance of proteins increases with BMI. The tables below show the
significant proteins for each comparison.

```{r table-sig-BMI1, results = "asis", echo = FALSE}
sig_tab <- significant_table(results, "BMIcat1")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_BMIcat1vs2.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for BMI cat 1 vs 2."))
```

```{r table-sig-BMI3, results = "asis", echo = FALSE}
sig_tab <- significant_table(results, "BMIcat3")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_BMIcat3vs2.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for BMI cat 3 vs 2."))
```

```{r table-sig-BMI4, results = "asis", echo = FALSE}
sig_tab <- significant_table(results, "BMIcat4")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_BMIcat4vs2.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for BMI cat 4 vs 2."))
```

An overlap of significant proteins can be seen for the comparisons all levels
(P01023 or Alpha-2-macroglobulin). Not unexpectedly, also the coefficient for
these proteins are larger in the comparison for BMI 4 than for BMI 3.

**A2M** (alpha-2-macroglobulin) is a protease inhibitor and cytokine
transporter. It can  inhibit inflammatory cytokines, and it thus disrupts
inflammatory cascades. A negative correlation between A2M mRNA expression and
BMI was described previously
[PMID:32927694](https://pubmed.ncbi.nlm.nih.gov/32927694/).

**C3** (complement C3) plays a central role in the activation of the complement
system. Its activation is required for both classical and alternative complement
activation pathways.

**CFH** (complement factor H) has an essential role in the regulation of
complement activation, restricting this innate defense mechanism to microbial
infections.

**APOA4** (apolipoprotein A4) encodes apolipoprotein which precise function is
not known. APOA4 is a potent activator of lecithin-cholesterol acetyltransferase
in vitro.

All 5 proteins according to tests p-values are visualized in boxplots in next
figures.
Points are shadowed according to sample age.

```{r, BP_BMI_1_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
bmi_1 <- prot_plot(prot_data, aes(x = BMIcat, y = P01024),
                   protein = "P01024")
ggplotly_or_not(bmi_1, PLOTLY)
```

```{r, BP_BMI_2_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
bmi_2 <- prot_plot(prot_data, aes(x = BMIcat, y = P05090),
                   protein = "P05090")
ggplotly_or_not(bmi_2, PLOTLY)
```

```{r, BP_BMI_3_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
bmi_3 <- prot_plot(prot_data, aes(x = BMIcat, y = P08603),
                   protein = "P08603")
ggplotly_or_not(bmi_3, PLOTLY)
```

```{r, BP_BMI_4_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
bmi_4 <- prot_plot(prot_data, aes(x = BMIcat, y = P01023),
                   protein = "P01023")
ggplotly_or_not(bmi_4, PLOTLY)
```

```{r, BP_BMI_5_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
bmi_4 <- prot_plot(prot_data, aes(x = BMIcat, y = P06727),
                   protein = "P06727")
ggplotly_or_not(bmi_4, PLOTLY)
```


### Identification of sex hormone related proteins

To identify sex hormone-related proteins, results from linear models fitted
previously are used.

```{r FIG-volcano-sexHorm, fig.path = IMAGE_PATH, fig.width = 8, fig.height = 8, fig.cap = "Volcano plot showing sex hormone-related associations. Red colored points are those with an adjusted p-value smaller than 0.05.", echo = FALSE, results = "hide"}
plot_volcano(results, "Sex_hormYes", main = "Sex_horm")
abline(h = -log10(0.05), lty = 2)
significant_points(cbind(results, sign_cv), comparison = "Sex_hormYes")
```

```{r table-sig-sex-horm, echo = FALSE, results = "asis"}
sig_tab <- significant_table(results, "Sex_hormYes")
## Exporting results
write_xlsx(sig_tab, path = paste0(dr, "/significant_Sex_horm.xlsx"))
pandoc.table(
    sig_tab,
    style = "rmarkdown", split.table = Inf,
    caption = paste0("Significant proteins for Sex horm. "))
```

**AGT** (angiotensinogen) is expressed in the liver and is cleaved by the enzyme
renin in response to lowered blood pressure. The protein is involved in
maintaining blood pressure, body fluid and electrolyte homeostasis.

**CP** (ceruloplasmin) is a metalloprotein that binds most of the copper in
plasma and is involved in the peroxidation of Fe(II)transferrin to
FE(III)transferrin.

**SHBG** (sex hormone binding globulin) encodes a protein which transports
androgens and estrogens in the blood.

**FETUB** (fetuin B) is part of a protein family which has been implicated in
osteogenesis, bone resorption, regulation of the insulin and hepatocyte growth
factor receptors, and response to systemic inflammation.

**SERPINA6** (serpin family A member 6) encodes an alpha-globulin protein with
corticosteroid-binding properties. This is the major transport protein for
glucocorticoids and progestins in the blood of most vertebrates.

**SERPINA7** (serpin family A member 7) encodes the major thyroid hormone
transport protein.

**SERPING1** (serpin family G member 1) is a highly glycosylated plasma protein
involved in the regulation of the complement cascade.

**PGLYRP2** (peptidoglycan recognition protein 2).

Top 5 proteins according to tests p-values are visualized in boxplots in next
figures.


```{r, BP_Sex_horm_1_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
sh_1 <- prot_plot(prot_data, aes(x = Sex_horm, y = P01019),
                  protein = "P01019")
ggplotly_or_not(sh_1, PLOTLY)
```

```{r, BP_Sex_horm_2_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
sh_2 <- prot_plot(prot_data, aes(x = Sex_horm, y = P00450),
                  protein = "P00450")
ggplotly_or_not(sh_2, PLOTLY)
```

```{r, BP_Sex_horm_3_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
sh_3 <- prot_plot(prot_data, aes(x = Sex_horm, y = Q96PD5),
                   protein = "Q96PD5")
ggplotly_or_not(sh_3, PLOTLY)
```

```{r, BP_Sex_horm_4_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
sh_4 <- prot_plot(prot_data, aes(x = Sex_horm, y = P04278),
                   protein = "P04278")
ggplotly_or_not(sh_4, PLOTLY)
```

```{r, BP_Sex_horm_5_figure, fig.path = IMAGE_PATH, fig.width=10, fig.height=6}
sh_5 <- prot_plot(prot_data, aes(x = Sex_horm, y = P08185),
                   protein = "P08185")
ggplotly_or_not(sh_5, PLOTLY)
```


# Sensitivity analysis

We next conduct a sensitivity analysis to evaluate the influence of hormonal
contraceptives on the results of potentially related traits such as sex or
age. To this end we fit linear regression models to the data that do not include
a variable for hormonal contraceptive use.

```{r}
FunForLinReg2 <- function(AnalyteName, Data) {
    Analyte <- Data[, AnalyteName]
    LMReg <- lm(Analyte ~ Sex + Age + Fasting + BMIcat,
                data = Data)
    res <- coef(summary(LMReg))[-1, ]
    p.Values <- data.frame(c(res[, 1],
                             res[, 4]))
    names(p.Values) <- AnalyteName
    rownames(p.Values) <- c(paste0("coef_", rownames(res)),
                            paste0("p-value_", rownames(res)))
    return(p.Values)
}
## do it in one run
LMRegtests <- do.call(
    cbind, lapply(proteins, function(x) FunForLinReg2(x, prot_data)))
Test_output_sens <- t(LMRegtests)
```

## Influence on sex-associated proteins

We compare the (raw) p-values and coefficients for association with sex from the
models with and without adjustment for use of hormonal contraceptives.

```{r sensitivity-sex, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 6, fig.cap = "Comparison of p-values (left) and coefficients (right) for association with sex from the models with (x-axis) and without adjustment for hormonal contraceptive use."}
par(mfrow = c(1, 2))
x <- Test_output[, "p-value_SexFemale"]
y <- Test_output_sens[, "p-value_SexFemale"]
x[x == 0] <- min(x[x > 0]) / 10
y[y == 0] <- min(y[y > 0]) / 10
x <- -log10(x)
y <- -log10(y)
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(log[10]~p-value[adj]),
     ylab = expression(log[10]~p-value),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "Sex association")
grid()
abline(0, 1)
x <- Test_output[, "coef_SexFemale"]
y <- Test_output_sens[, "coef_SexFemale"]
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(coef[adj]),
     ylab = expression(coef),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "Sex association")
grid()
abline(0, 1)
```

Coefficients are larger and p-values smaller for proteins without adjustment for
hormonal contraceptive usage.


## Influence on age-associated proteins

We compare the (raw) p-values and coefficients for association with sex from the
models with and without adjustment for use of hormonal contraceptives.

```{r sensitivity-age, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 6, fig.cap = "Comparison of p-values (left) and coefficients (right) for association with age from the models with (x-axis) and without adjustment for hormonal contraceptive use."}
par(mfrow = c(1, 2))
x <- Test_output[, "p-value_Age"]
y <- Test_output_sens[, "p-value_Age"]
x[x == 0] <- min(x[x > 0]) / 10
y[y == 0] <- min(y[y > 0]) / 10
x <- -log10(x)
y <- -log10(y)
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(log[10]~p-value[adj]),
     ylab = expression(log[10]~p-value),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "Age association")
grid()
abline(0, 1)
x <- Test_output[, "coef_Age"]
y <- Test_output_sens[, "coef_Age"]
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(coef[adj]),
     ylab = expression(coef),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "Age association")
grid()
abline(0, 1)
```

The effect of the adjustment is even bigger for age associated proteins.


## Influence on BMI-associated proteins

We compare the (raw) p-values and coefficients for association with BMI
(category 4) from the models with and without adjustment for use of hormonal
contraceptives.

```{r sensitivity-bmi4, fig.path = IMAGE_PATH, fig.width = 12, fig.height = 6, fig.cap = "Comparison of p-values (left) and coefficients (right) for association with BMI from the models with (x-axis) and without adjustment for hormonal contraceptive use."}
par(mfrow = c(1, 2))
x <- Test_output[, "p-value_BMIcat4"]
y <- Test_output_sens[, "p-value_BMIcat4"]
x[x == 0] <- min(x[x > 0]) / 10
y[y == 0] <- min(y[y > 0]) / 10
x <- -log10(x)
y <- -log10(y)
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(log[10]~p-value[adj]),
     ylab = expression(log[10]~p-value),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "BMI 4 vs 2")
grid()
abline(0, 1)
x <- Test_output[, "coef_BMIcat4"]
y <- Test_output_sens[, "coef_BMIcat4"]
xl <- range(c(x, y))
plot(x = x, y = y, xlim = xl, ylim = xl,
     xlab = expression(coef[adj]),
     ylab = expression(coef),
     pch = 21, col = "#00000080", bg = "#00000040",
     main = "BMI 4 vs 2")
grid()
abline(0, 1)
```

Adjustment for hormonal contraceptive use has only a minimal influence on the
results for BMI.

# Data, figures etc for the manuscript's supplement

```{r}
dist_plot <- function(x, protein, ...) {
    plot(density(x[, protein]), xlab = expression(log[2]~abundance),
         main = paste0(protein, " (", prot_ann[protein, "Genes"], ")"), ...)
    grid()
}

dr <- paste0("images/manuscript/supplement/")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)
for (protein in proteins) {
    fl <- paste0(dr, "density-", protein, ".png")
    png(fl, width = 10, height = 4, units = "cm", res = 300, pointsize = 4)
    par(mar = c(4.5, 4.2, 1.5, 0.5))
    dist_plot(prot_data, protein, lwd = 2)
    dev.off()
}
```


# Correlation between clinical/lab parameters and plasma proteomics data

We now consider selected clinical laboratory parameters measured for CHRIS
participants and perform a correlation analysis with related plasma protein
concentrations. The table below lists the laboratory parameters and selected
protein as well as their correlation coefficient (Spearman rho). Since most
laboratory measurements don't quantify directly proteins we don't expect a
perfect similarity between the mass spectrometry and laboratory assay based
measurements.

```{r, results = "asis"}
pairs <- data.frame(
    label = c("x0lp04q", "x0lp09", "x0lp17", "x0lp18",
              "x0lp19", "x0lp29", "x0lp41"),
    Uniprot = c("P01008", "P02768", "P02647", "P04114",
                "P02656", "P02787", "P69905"))
anns <- chrisTraitAnnotations()

pairs$parameter_name <- anns[pairs$label, "name"]
pairs$hgnc_symbol <- prot_ann[pairs$Uniprot, "Genes"]
pairs$description <- prot_ann[pairs$Uniprot, "long_description"]

pairs$rho <- NA_real_
for (i in seq_len(nrow(pairs))) {
    pairs$rho[i] <- cor(prot_data[, pairs[i, "Uniprot"]],
                        labdata_info[rownames(prot_data), pairs[i, "label"]],
                        use = "pairwise.complete.obs", method = "spearman")
}
pandoc.table(pairs, style = "rmarkdown", split.tables = Inf,
             caption = "Correlation with selected laboratory parameters")
```

Apolipoprotein A-I is the major structural component of HDL while apolipoprotein
B is the primary apolipoprotein of LDL

```{r}
for (i in seq_len(nrow(pairs))) {
    plot(labdata_info[rownames(prot_data), pairs[i, "label"]],
         2^prot_data[, pairs[i, "Uniprot"]],
         pch = 21, col = "#00000080", bg = "#00000040",
         xlab = pairs[i, "parameter_name"], ylab = pairs[i, "hgnc_symbol"])
    grid()
}
```

- *x0lp04q* (Antithrombin (%)): measured in plasma citrate using the enzymatic
  Siemens Innovance Anthithrombine assay on a ROCHE SYSMEX CA1500 to quantify
  the functionally active antithrombin; after 2014-01 using the enzymatic
  STA-Stachrom AT III - REF 00596 assay on a STAGO STA COMPACT MAX instrument.
- *x0lp09* (Albumin (g/dL): measured in serum using the colorimetric Cobas ALB
  plus Albumin BCG assay on a ROCHE MODULAR PPE and since 2014-05 using the
  colorimetric ALBUMIN BCG assay on an ABBOT DIAGNOSTIC ARCHITECT instrument.
- *x0lp17* (HDL (mg/dL)): measured in serum using the enzymatic Cobas HDL-C PLUS
  3rd generation assay on a ROCHE MODULAR PPE instrument. After 2014-05 using
  the ULTRA HDL on an ABBOT DIAGNOSTIC ARCHITECT instrument.
- *x0lp18* (LDL (mg/dL)): measured in serum using the enzymatic Cobas LDL-C plus
  2nd generation assay on a ROCHE MODULAR PPE and since 2014-05 using the DIRECT
  LDL assay on an ABBOT DIAGNOSTIC ARCHITECT instrument.
- *x0lp19* (Triglycerides (mg/dL))): measured in serum using the enzymatic Cobas
  Triglyceride GPO-PAP assay on a ROCHE MODULAR PPE and since 2014-05 using the
  TRIGLYCERIDE assay on an ABBOT DIAGNOSTIC ARCHITECT instrument.
- *x0lp29* (Transferrin (mg/dL)): measured in serum using the immunological
  Cobas Tina-quant Transferring ver.2 assay on an ROCHE MODULAR PPE
  instrument. Since 2014-05 using the immunological TRANSFERRIN assay on an
  ABBOT DIAGNOSTIC ARCHITECT.
- *x0lp41* (HGB (g/dL)): measured in EDTA plasma using the electronic impedance
  laser light scattering based assay with Cell-Dyn Sapphire reagents on an ABBOT
  CD SAPPHIRE instrument and since 2014-03 on a SYSMEX XN-1000 (also electronic
  impedance laser light scattering based assay).

```{r}
png("images/manuscript/Figure_1.png", width = 10, height = 5, units = "cm",
    res = 600, pointsize = 4)

par(mfrow = c(1, 2), mar = c(4.5, 4.3, 0.5, 0.5), cex.lab = 1.5, bty = "n")
plot(labdata_info[rownames(prot_data), pairs[4, "label"]],
     2^prot_data[, pairs[4, "Uniprot"]],
     pch = 16, col = "#00000040",
     xlab = pairs[4, "parameter_name"], ylab = pairs[4, "hgnc_symbol"])
grid()
plot(labdata_info[rownames(prot_data), pairs[6, "label"]],
     2^prot_data[, pairs[6, "Uniprot"]],
     pch = 16, col = "#00000040",
     xlab = pairs[6, "parameter_name"], ylab = pairs[6, "hgnc_symbol"])
grid()
dev.off()

```



# Session information

R packages used for the analysis:

```{r}
sessionInfo()
```


# References
