---
title: "Supplementary Information for: \\textit{...}"
output:
  pdf_document:
    includes:
      in_header: "header.sty"
    toc: true
    toc_depth: 2
---

\setcounter{table}{3}

```{r style, echo = FALSE, results = 'asis', message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# General data overview

## Signal distributions of individual proteins

Distribution of $log_{2}$ transformed signal distributions in the analyzed
samples is shown for each protein.

```{r, results = "asis"}
fls <- dir("images/manuscript/supplement", pattern = "^density",
           full.names = TRUE)

for (i in seq_along(fls)) {
    prot_id <- sub("^density-", "", basename(fls[i]))
    prot_id <- sub(".png", "", prot_id, fixed = TRUE)
    cat("\n\\begin{figure}[H]\n")
    cat("\\centering\n")
    cat("  \\includegraphics[width=0.7\\textwidth]{",
        fls[i], "}\n", sep = "")
    cat("  \\caption{Signal distribution for ", prot_id, ".}\n", sep = "")
    cat("\\end{figure}\n")
    ## cat(paste0("![", met_name[i],".](", met_image[i], ")"), "\n")
    }
```


## Principal Component Analysis and influence of hormonal contraceptive use

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.7\\textwidth]{",
    "images/manuscript/supplement/Figure_S149.png}\n",
    sep = "")
cat("  \\caption{Principal component anaysis of the plasma proteome data. ",
    "Data points are colored according to usage of hormonal contraceptives.}\n",
    sep = "")
cat("\\end{figure}\n")

```

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.7\\textwidth]{",
    "images/manuscript/supplement/Figure_S150.png}\n", sep = "")
cat("  \\caption{Abundance of the protein Angiotensinogen (AGT) in study ",
    "participants using hormonal contraceptives (HCU) and female and male ",
    "participants that don't.}\n",
    sep = "")
cat("\\end{figure}\n")

```


# Sex, age, BMI and hormonal contraceptive use-associated proteins

## Sex-associated plasma proteins

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
library(readxl)
library(pander)
results <- as.data.frame(read_xlsx(
    paste0("data/xlsx/sex_age_bmi_associated_proteins/",
           "results_sex_age_bmi_sex_hormone_adjusted.xlsx")))
rownames(results) <- results$Uniprot

tab <- results[results$significant_SexFemale,
               c("Description", "Genes",
                 "coef_SexFemale",
                 "p.adj_SexFemale",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_SexFemale), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between female ",
        "and male study participants. $coef$ and $p_{adj}$: coefficient ",
        "(representing the differential abundance in log2 scale) and p-value ",
        "adjusted for multiple hypothesis testing. $A_{F}$ and $A_{M}$:",
        " average abundance in female and male participants, respectively. ",
        "Proteins are ordered by p-value."))

```


## Age-associated plasma proteins

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_Age,
               c("Description", "Genes",
                 "coef_Age",
                 "p.adj_Age",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_Age), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins significantly associated with participants' age. ",
        "$coef$ and $p_{adj}$: coefficient (representing the change in ",
        "abundance per year) and p-value adjusted for multiple hypothesis ",
        "testing. $A_{F}$ and $A_{M}$: average abundance in female and ",
        "male participants, respectively. ",
        "Proteins are ordered by p-value."))
```

## Body mass index associated plasma proteins

Proteins with significant differences in concentrations between body mass
index (BMI) categories 1 (BMI < 18.5), 3 (25 <= BMI < 30) and 4 (BMI > 30) to
the *normal* category 2 (18.5 <= BMI < 25) are listed in the tables
below. Proteins with an adjusted p-value smaller than 0.05 and a difference
in concentrations larger than the coefficient of variation in QC (study
pool) samples are considered significant.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat1,
               c("Description", "Genes",
                 "coef_BMIcat1",
                 "p.adj_BMIcat1",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_BMIcat1), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI < 18.5 (BMI1) and participants with a BMI ",
        "between 18.5 and 25 (BMI2; normal). ",
        "$coef$ and $p_{adj}$: coefficient (representing the differential ",
        "abundance in log2 scale) and p-value adjusted for multiple ",
        "hypothesis testing. $A_{F}$ and $A_{M}$:",
        " average abundance in female and male participants, respectively. ",
        "Proteins are ordered by p-value."))

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat3,
               c("Description", "Genes",
                 "coef_BMIcat3",
                 "p.adj_BMIcat3",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_BMIcat3), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI between 25 and 30 (BMI3) and participants ",
        "with a BMI between 18.5 and 25 (BMI2; normal). ",
        "$coef$ and $p_{adj}$: coefficient (representing the differential ",
        "abundance in log2 scale) and p-value adjusted for multiple ",
        "hypothesis testing. $A_{F}$ and $A_{M}$:",
        " average abundance in female and male participants, respectively. ",
        "Proteins are ordered by p-value."))

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat4,
               c("Description", "Genes",
                 "coef_BMIcat4",
                 "p.adj_BMIcat4",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_BMIcat4), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI > 30 (BMI4) and participants with a BMI ",
        "between 18.5 and 25 (BMI2; normal). ",
        "$coef$ and $p_{adj}$: coefficient (representing the differential ",
        "abundance in log2 scale) and p-value adjusted for multiple ",
        "hypothesis testing. $A_{F}$ and $A_{M}$:",
        " average abundance in female and male participants, respectively. ",
        "Proteins are ordered by p-value."))

```


## Plasma proteins associated with usage of hormonal contraceptives

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_Sex_hormYes,
               c("Description", "Genes",
                 "coef_Sex_hormYes",
                 "p.adj_Sex_hormYes",
                 "Avg_Female",
                 "Avg_Male")]
tab <- tab[order(tab$p.adj_Sex_hormYes), ]
colnames(tab) <- c("$Description$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$A_{F}$", "$A_{M}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccllll", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between study ",
        "participants that use hormonal contraceptives and participants that ",
        "don't.",
        " $coef$ and $p_{adj}$: coefficient (representing the differential ",
        "abundance in log2 scale) and p-value ",
        "adjusted for multiple hypothesis testing. $A_{F}$ and $A_{M}$:",
        " average abundance in female and male participants, respectively. ",
        "Proteins are ordered by p-value."))

```
