---
title: "Supplementary Information for: \\textit{Hormonal Contraceptives are Shaping the Human Plasma Proteome in a Large Population Cohort}"
output:
  pdf_document:
    includes:
      in_header: "header.sty"
    toc: true
    toc_depth: 2
---

\setcounter{table}{2}

```{r style, echo = FALSE, results = 'asis', message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# General data overview

## Signal distributions of individual proteins

Distribution of $log_{2}$ transformed signal distributions in the analyzed
samples is shown for each protein.

```{r, results = "asis"}
library(readxl)
library(pander)
results <- as.data.frame(read_xlsx(
    paste0("data/xlsx/sex_age_bmi_associated_proteins/",
           "results_sex_age_bmi_hcu_adjusted.xlsx")))
rownames(results) <- results$Uniprot

fls <- dir("images/manuscript/supplement", pattern = "^density",
           full.names = TRUE)

for (i in seq_along(fls)) {
    prot_id <- sub("^density-", "", basename(fls[i]))
    prot_id <- sub(".png", "", prot_id, fixed = TRUE)
    cat("\n\\begin{figure}[H]\n")
    cat("\\centering\n")
    cat("  \\includegraphics[width=0.7\\textwidth]{",
        fls[i], "}\n", sep = "")
    cat("  \\caption{Signal distribution for ", prot_id,
        " (", results[prot_id, "Genes"],").}\n", sep = "")
    cat("\\end{figure}\n")
    ## cat(paste0("![", met_name[i],".](", met_image[i], ")"), "\n")
    }
```

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.7\\textwidth]{",
    "images/manuscript/supplement/Figure_CV.png}\n",
    sep = "")
cat("  \\caption{Distribution of coefficients of variation (CV, in \\%) before",
    " and after data pre-processing. Left: CVs of raw peptide data ",
    "(DIA-NN normalized data). Right: CVs of protein data after ",
    "normalization}\n",
    sep = "")
cat("\\end{figure}\n")

```

**Table S1** (external spreadsheet, xlsx format): Results from association
analyses.

**Table S2** (external spreadsheet, xlsx format): Functional annotations for
detected proteins.

```{r, results = "asis"}
tab <- results[,
               c("Uniprot", "Genes",
                 "cv_study",
                 "cv_qc_chris",
                 "relative_cv"
                 )]
colnames(tab) <- c("$UniProt$", "$Gene$", "$CV_{study}$", "$CV_{QC}$",
                   "$CV_{rel}$")
rownames(tab) <- NULL
tab_sub <- tab[order(tab[, 3]), ][1:30, ]
rownames(tab_sub) <- NULL
pandoc.table(
    tab_sub, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with lowest coefficient of variation across study samples. ",
        "$CV_{study}$: coefficient of variation in study samples, ",
        "$CV_{QC}$: coefficient of variation in pooled QC samples and ",
        "$CV_{rel}$: relative coefficient of variation expressed as the ratio ",
        "between the CV in study and in pooled QC samples."))

```

```{r, results = "asis"}
tab_sub <- tab[order(tab[, 5], decreasing = TRUE), ][1:30, ]
rownames(tab_sub) <- NULL
pandoc.table(
    tab_sub, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with the highest (relative) coefficient of variation across ",
        "study samples. ",
        "$CV_{study}$: coefficient of variation in study samples, ",
        "$CV_{QC}$: coefficient of variation in pooled QC samples and ",
        "$CV_{rel}$: relative coefficient of variation expressed as the ratio ",
        "between the CV in study and in pooled QC samples."))

```

```{r, results = "asis"}
## FDA Biomarkers
fda <- read_xlsx("data/xlsx/FDA_approved_biomarkers.xlsx") |> as.data.frame()
fda <- fda[!is.na(fda$uniprot_id), ]
tab <- results[fda$uniprot_id, c("Uniprot", "Genes", "cv_study", "cv_qc_chris")]
colnames(tab) <- c("$UniProt$", "$Genes$", "$CV_{study}$", "$CV_{QC}$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Common disease biomarker proteins quantified in the present data set. ",
        "$CV_{study}$: coefficient of variation in study samples and ",
        "$CV_{QC}$: coefficient of variation in pooled QC samples."))

```

```{r, results = "asis"}
tmp <- data.frame(
    `Laboratory parameter` = c("Antithrombin (%)", "Albumin (g/dL)",
                               "HDL (mg/dL)", "LDL (mg/dL)",
                               "Triglycerides (mg/dL)", "Transferrin (mg/dL)",
                               "HGN (h/dL)"),
    `UniProt ID` = c("P01008", "P02768", "P02647", "P04114", "P02656",
                     "P02787", "P69905"),
    `HGNC symbol` = c("SERPINC1", "ALB", "APOA1", "APOB", "APOC3",
                      "TF", "HBA1"),
    rho = c(0.366, 0.320, 0.408, 0.787, 0.596, 0.686, 0.480))
pandoc.table(
    tmp, split.table = "Inf", justity = c("cccc"),
    caption = paste0("Correlation between clinical laboratory measurements",
                     " and quantified protein abundances."))
```

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.7\\textwidth]{",
    "images/manuscript/supplement/Figure_1.png}\n",
    sep = "")
cat("  \\caption{Correlation between quantified protein abundances and ",
    "related diagnostic assays. Shown are MS-determind, relative protein ",
    "abundances (normalized signal intensities) on the y-axis and ",
    "related diagnostic assays on the x-axis.}\n",
    sep = "")
cat("\\end{figure}\n")

```


## Principal Component Analysis and influence of hormonal contraceptive use

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.7\\textwidth]{",
    "images/manuscript/supplement/Figure_S149.png}\n",
    sep = "")
cat("  \\caption{Principal component anaysis of the plasma proteome data. ",
    "Data points are colored according to usage of hormonal contraceptives.}\n",
    sep = "")
cat("\\end{figure}\n")

```



# Sex, age and body mass index associated proteins

## Sex-associated plasma proteins

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
library(readxl)
library(pander)
results <- as.data.frame(read_xlsx(
    paste0("data/xlsx/sex_age_bmi_associated_proteins/",
           "results_sex_age_bmi_hcu_adjusted.xlsx")))
rownames(results) <- results$Uniprot

tab <- results[results$significant_SexFemale,
               c("Uniprot", "Genes",
                 "coef_SexFemale",
                 "p.adj_SexFemale",
                 "effect_size_SexFemale"
                 )]
tab <- tab[order(tab$p.adj_SexFemale), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between female ",
        "and male study participants. $coef$, $ES$ and $p_{adj}$: coefficient ",
        "(representing the differential abundance in log2 scale), effect size",
        " and p-value adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))

```


## Age-associated plasma proteins

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_Age,
               c("Uniprot", "Genes",
                 "coef_Age_10",
                 "p.adj_Age_10",
                 "effect_size_Age_10"
                 )]
tab <- tab[order(tab$p.adj_Age_10), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins significantly associated with participants' age. ",
        "$coef$, $ES$ and $p_{adj}$: coefficient (representing the log2 change",
        " in abundance in 10 years), effect size (for 10-year change) and ",
        "p-value adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))
```

The age association for the 2 most significant proteins is shown below.

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.8\\textwidth]{",
    "images/manuscript/supplement/Figure_age_1.png}\n",
    sep = "")
cat("  \\caption{Age dependency of the protein IGFALS. Shown are sex, BMI, ",
    "HCU and fasting status adjusted abundances against participants' age. ",
    "The solid black line represents the linear regression line..}\n",
    sep = "")
cat("\\end{figure}\n")

```

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.8\\textwidth]{",
    "images/manuscript/supplement/Figure_age_2.png}\n",
    sep = "")
cat("  \\caption{Age dependency of the protein VTN. Shown are sex, BMI, ",
    "HCU and fasting status adjusted abundances against participants' age. ",
    "The solid black line represents the linear regression line..}\n",
    sep = "")
cat("\\end{figure}\n")

```


## Body mass index associated plasma proteins

Proteins with significant differences in concentrations between body mass
index (BMI) categories 1 (BMI < 18.5), 3 (25 <= BMI < 30) and 4 (BMI > 30) to
the *normal* category 2 (18.5 <= BMI < 25) are listed in the tables
below. Proteins with an adjusted p-value smaller than 0.05 and a difference
in concentrations larger than the coefficient of variation in QC (study
pool) samples are considered significant.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat1,
               c("Uniprot", "Genes",
                 "coef_BMIcat1",
                 "p.adj_BMIcat1",
                 "effect_size_BMIcat1"
                 )]
tab <- tab[order(tab$p.adj_BMIcat1), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI < 18.5 (BMI1) and participants with a BMI ",
        "between 18.5 and 25 (BMI2; normal). ",
        "$coef$, $ES$ and $p_{adj}$: coefficient (representing the ",
        "differential abundance in log2 scale), effect size and p-value ",
        "adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat3,
               c("Uniprot", "Genes",
                 "coef_BMIcat3",
                 "p.adj_BMIcat3",
                 "effect_size_BMIcat3"
                 )]
tab <- tab[order(tab$p.adj_BMIcat3), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI between 25 and 30 (BMI3) and participants ",
        "with a BMI between 18.5 and 25 (BMI2; normal). ",
        "$coef$, $ES$ and $p_{adj}$: coefficient (representing the ",
        "differential abundance in log2 scale) effect size and p-value ",
        "adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_BMIcat4,
               c("Uniprot", "Genes",
                 "coef_BMIcat4",
                 "p.adj_BMIcat4",
                 "effect_size_BMIcat4"
                 )]
tab <- tab[order(tab$p.adj_BMIcat4), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants with a BMI > 30 (BMI4) and participants with a BMI ",
        "between 18.5 and 25 (BMI2; normal). ",
        "$coef$, $ES$ and $p_{adj}$: coefficient (representing the ",
        "differential abundance in log2 scale) effect size and p-value ",
        "adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))

```


## Proteins significantly associated with fasting status

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
tab <- results[results$significant_FastingNo,
               c("Uniprot", "Genes",
                 "coef_FastingNo",
                 "p.adj_FastingNo",
                 "effect_size_FastingNo"
                 )]
tab <- tab[order(tab$p.adj_FastingNo), ]
colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                   "$ES$")
rownames(tab) <- NULL
pandoc.table(
    tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins with significant difference in abundance between ",
        "participants declared to not have fasted against those who did. ",
        "$coef$, $ES$ and $p_{adj}$: coefficient (representing the ",
        "differential abundance in log2 scale) effect size and p-value ",
        "adjusted for multiple hypothesis testing. ",
        "Proteins are ordered by p-value."))

```




# Plasma proteins associated with usage of hormonal contraceptives


A comparison of the results from the analysis on the full data set and on the
subset of female participants below 40 years of age is shown below. Coefficients
for associations with hormonal contraceptive use are highly similar between the
two analyses and p-values highly related (same rank, but difference in the
values due to the differences in statistical power from the two analyses).

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.9\\textwidth]{",
    "images/manuscript/supplement/Figure_S151.png}\n",
    sep = "")
cat("  \\caption{Comparison of results for association with hormonal ",
    "contraceptive use from the analysis of the full data set (x-axis) and ",
    "the analysis on the subset of female participants below 40 years ",
    "of age (y axis). Shown are -log10 of p-values, ranks of p-values and ",
    "coefficients.}\n",
    sep = "")
cat("\\end{figure}\n")

```


## Sensitivity analysis

To evaluate the influence of hormonal contraceptive use on results for age, sex
and BMI associations, the analysis was repeated without adjustment for hormonal
contraceptive use. Proteins that were found significantly associated with sex in
this analysis, but were no longer significant after adjustment for hormonal
contraceptive use are listed in the table below.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
a <- read_xlsx(
    "data/xlsx/sex_age_bmi_associated_proteins/sensitivity_sex.xlsx") |>
as.data.frame()

colnames(a) <- c("$Gene$", "$coef$", "$p$", "$coef_{adj}$",
                 "$p_{adj}$", "$coef_{HCU}$", "$p_{HCU}$")
rownames(a) <- NULL
pandoc.table(
    a, split.table = "Inf", justify = "crrrrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins significantly associated with sex in an analysis without ",
        "adjusting for hormonal contraceptive use that were no longer ",
        "significant after adjustment. $coef$ and $p$: coefficient and ",
        "bonferroni adjusted p-value from the analysis without adjustment for ",
        "hormonal contraceptive use. $coef_{adj}$, $p_{adj}$, $coef_{HCU}$ ",
        "and $p_{HCU}$: coefficients and bonferroni adjusted p-values for ",
        "sex association and hormoman contraceptive use association from ",
        "the analysis with adjustment for hormonal contraceptive use."))

```

Proteins significantly associated with age from an analysis without adjustment
for hormonal contraceptive use that were no longer significant after adjustment
for hormonal contraceptive use are listed in the table below.

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "asis"}
b <- read_xlsx(
    "data/xlsx/sex_age_bmi_associated_proteins/sensitivity_age.xlsx") |>
as.data.frame()

colnames(b) <- c("$Gene$", "$coef$", "$p$", "$coef_{adj}$",
                 "$p_{adj}$", "$coef_{HCU}$", "$p_{HCU}$")
rownames(b) <- NULL
pandoc.table(
    b, split.table = "Inf", justify = "crrrrrr", keep.line.breaks = FALSE,
    split.cells = 50, style = "rmarkdown", digits = 3,
    caption = paste0(
        "Proteins significantly associated with age in an analysis without ",
        "adjusting for hormonal contraceptive use that were no longer ",
        "significant after adjustment. $coef$ and $p$: coefficient and ",
        "bonferroni adjusted p-value from the analysis without adjustment for ",
        "hormonal contraceptive use. $coef_{adj}$, $p_{adj}$, $coef_{HCU}$ ",
        "and $p_{HCU}$: coefficients and bonferroni adjusted p-values for ",
        "age association and hormonal contraceptive use association from ",
        "the analysis with adjustment for hormonal contraceptive use."))

```

# Plasma proteins associated with medication (ATC level 3)

Associations were evaluated for 27 ATC level 3 medications taken on a regular
basis (at least twice per week) by more than 14 study participants. Significant
associations are shown for each medication in the tables below. For 11
medications no significant association was identified.

```{r, results = "asis"}
atc3 <- read_xlsx("data/xlsx/sex_age_bmi_associated_proteins/results_sex_age_bmi_atc3.xlsx")
atc3 <- as.data.frame(atc3)

atc_3_names <- rbind(
    c("G03A", "HORMONAL CONTRACEPTIVES FOR SYSTEMIC USE",
      ".HORMONAL.CONTRACEPTIVES.FOR.SYSTEMIC.USE.Yes"),
    c("G03H", "ANTIANDROGENS", "ANTIANDROGENSYes"),
    c("G02B", "CONTRACEPTIVES FOR TOPICAL USE",
      ".CONTRACEPTIVES.FOR.TOPICAL.USE.Yes"),
    c("M01A", "ANTIINFLAMMATORY AND ANTIRHEUMATIC PRODUCTS, NON-STEROIDS",
      ".ANTIINFLAMMATORY.AND.ANTIRHEUMATIC.PRODUCTS..NON.STEROIDS.Yes"),
    c("N04B", "DOPAMINERGIC AGENTS", ".DOPAMINERGIC.AGENTS.Yes"),
    c("A10B", "BLOOD GLUCOSE LOWERING DRUGS, EXCL. INSULINS",
      ".BLOOD.GLUCOSE.LOWERING.DRUGS..EXCL..INSULINS.Yes"),
    c("M04A", "ANTIGOUT PREPARATIONS", ".ANTIGOUT.PREPARATIONS.Yes"),
    c("B01A", "ANTITHROMBOTIC AGENTS", ".ANTITHROMBOTIC.AGENTS.Yes"),
    c("C07A", "BETA BLOCKING AGENTS", ".BETA.BLOCKING.AGENTS.Yes"),
    c("G04C", "DRUGS USED IN BENIGN PROSTATIC HYPERTROPHY",
      ".DRUGS.USED.IN.BENIGN.PROSTATIC.HYPERTROPHY.Yes"),
    c("N03A", "ANTIEPILEPTICS", "ANTIEPILEPTICSYes"),
    c("R03A", "ADRENERGICS, INHALANTS", ".ADRENERGICS..INHALANTS.Yes"),
    c("A12A", "CALCIUM", "CALCIUMYes"),
    c("C10A", "LIPID MODIFYING AGENTS, PLAIN",
      ".LIPID.MODIFYING.AGENTS..PLAIN.Yes"),
    c("C09D", "ANGIOTENSIN II ANTAGONISTS, COMBINATIONS",
      ".ANGIOTENSIN.II.ANTAGONISTS..COMBINATIONS.Yes"),
    c("C09C", "ANGIOTENSIN II ANTAGONISTS, PLAIN",
      ".ANGIOTENSIN.II.ANTAGONISTS..PLAIN.Yes")
) |> as.data.frame()
colnames(atc_3_names) <- c("atc", "name", "var")


for (i in seq_len(nrow(atc_3_names))) {
    tab <- atc3[atc3[, paste0("significant_", atc_3_names$var[i])],
                c("Uniprot", "Genes",
                  paste0("coef_", atc_3_names$var[i]),
                  paste0("p.adj_", atc_3_names$var[i]),
                  paste0("effect_size_", atc_3_names$var[i]))]
    tab <- tab[order(tab[, paste0("p.adj_", atc_3_names$var[i])]), ]
    colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                       "$ES$")
    rownames(tab) <- NULL
    pandoc.table(
        tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
        split.cells = 50, style = "rmarkdown", digits = 3,
        caption = paste0(
            "Proteins significantly associated with ATC level 3 medication *",
            atc_3_names$name[i], "* (ATC3 ", atc_3_names$atc[i], ").",
            " $coef$ and $p_{adj}$: coefficient (representing the ",
            "differential abundance in log2 scale) and p-value ",
            "adjusted for multiple hypothesis testing. $ES$ effect size. ",
            "Proteins are ordered by p-value."))
}
```

# Plasma proteins associated with medication (ATC level 4)

Associations were evaluated for 27 ATC level 4 medications taken on a regular
basis (at least twice per week) by more than 14 study participants. The
considered medications, numbers of participants and significant proteins are
shown in the table below.

```{r, results = "asis"}
atc4_table <- data.frame(
    ATC = c("H03AA", "G03AA, G03FA", "B01AC", "C10AA", "C09AA", "C07AB",
            "C09DA", "N06AB", "A02BC", "C08CA", "C09CA", "A12AX", "C09BA",
            "N05BA, N05CD", "G04CA, C02CA", "A10BA", "N06AX", "B01AA",
            "M04AA", "G03AB, G03FB", "R03AK", "G02BA", "M01AE", "S01ED",
            "G02BB", "N03AX", "H02AB, R03BA", "M01AB", "G03HB"),
    Name = c("Thyroid hormones",
             "Progestogens and estrogens, fixed combinations",
             "Platelet aggregation inhibitors excl. heparin",
             "HMG CoA reductase inhibitors", "ACE inhibitors, plain",
             "Beta blocking agents, selective",
             "Angiotensin II antagonists and diuretics",
             "Selective serotonin reuptake inhibitors",
             "Proton pump inhibitors", "Dihydropyridine derivatives",
             "Angiotensin II antagonists, plain",
             "Calcium, combinations with vitamin D and/or other drugs",
             "ACE inhibitors and diuretics", "Benzodiazepine derivatives",
             "Alpha-adrenoreceptor antagonists", "Biguanides",
             "Other antidepressants", "Vitamin K antagonists",
             "Preparations inhibiting uric acid production",
             "Progestogens and estrogens, sequential preparations",
             "Adrenergics in combination with corticosteroids or other drugs, excl. anticholinergics",
             "Intrauterine contraceptives",
             "Propionic acid derivatives",
             "Beta blocking agents",
             "Intravaginal contraceptives",
             "Other antiepileptics",
             "Glucocorticoids",
             "Acetic acid derivatives and related substances",
             "Antiandrogens and estrogens"),
    Participants = c(262, 260, 211, 193, 156, 135, 109, 94, 92, 91, 74, 62,
                     59, 51, 48, 43, 41, 37, 36, 34, 27, 25, 24, 21, 21, 21,
                     18, 16, 16),
    Proteins = c(0, 52, 1, 1, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 5, 4, 0, 14, 0,
                 36, 1, 0, 0, 0, 35, 0, 2, 0, 38)
)
pandoc.table(
    atc4_table, split.table = "Inf", justify = "ccrr", keep.line.breaks = FALSE,
        split.cells = 50, style = "rmarkdown", digits = 3,
        caption = paste0("Overview of association results for ATC level 4 ",
                         "medications in the CHRIS study subset. Only ",
                         "medications taken on a regular basis by more than ",
                         "14 of the in total 3,632 study participants were ",
                         "considered. Columns *Participants* and *Proteins* ",
                         "list the number of participants taking the ",
                         "medication and number of significantly associated ",
                         "proteins."))
```

```{r, results = "asis"}
cat("\n\\begin{figure}[H]\n")
cat("\\centering\n")
cat("  \\includegraphics[width=0.9\\textwidth]{",
    "images/manuscript/supplement/Figure_ATC4.png}\n",
    sep = "")
cat("  \\caption{Significant associations between proteins (rows) and ATC",
    " level 4 medications (columns). Effect sizes (as a number and color coded)",
    " are only shown for significant associations.}\n",
    sep = "")
cat("\\end{figure}\n")

```

```{r, results = "asis"}
atc4 <- read_xlsx("data/xlsx/sex_age_bmi_associated_proteins/results_sex_age_bmi_atc4.xlsx")
atc4 <- as.data.frame(atc4)

atc4_table$var <- c(
    ".Thyroid.hormones.Yes",
    ".Progestogens.and.estrogens..fixed.combinations.Yes",
    ".Platelet.aggregation.inhibitors.excl..heparin.Yes",
    ".HMG.CoA.reductase.inhibitors.Yes",
    ".ACE.inhibitors..plain.Yes",
    ".Beta.blocking.agents..selective.Yes",
    ".Angiotensin.II.antagonists.and.diuretics.Yes",
    ".Selective.serotonin.reuptake.inhibitors.Yes",
    ".Proton.pump.inhibitors.Yes",
    ".Dihydropyridine.derivatives.Yes",
    ".Angiotensin.II.antagonists..plain.Yes",
    ".Calcium..combinations.with.vitamin.D.and.or.other.drugs.Yes",
    ".ACE.inhibitors.and.diuretics.Yes",
    ".Benzodiazepine.derivatives.Yes",
    ".Alpha.adrenoreceptor.antagonists.Yes",
    "BiguanidesYes",
    ".Other.antidepressants.Yes",
    ".Vitamin.K.antagonists.Yes",
    ".Preparations.inhibiting.uric.acid.production.Yes",
    ".Progestogens.and.estrogens..sequential.preparations.Yes",
    ".Adrenergics.in.combination.with.corticosteroids.or.other.drugs..excl..anticholinergics.Yes",
    ".Intrauterine.contraceptives.Yes",
    ".Propionic.acid.derivatives.Yes",
    ".Beta.blocking.agents.Yes",
    ".Intravaginal.contraceptives.Yes",
    ".Other.antiepileptics.Yes",
    "GlucocorticoidsYes",
    ".Acetic.acid.derivatives.and.related.substances.Yes",
    ".Antiandrogens.and.estrogens.Yes")

atc4_table <- atc4_table[atc4_table$Proteins > 0, ]
atc4_table <- atc4_table[order(atc4_table$Proteins, decreasing = TRUE), ]

for (i in seq_len(nrow(atc4_table))) {
    tab <- atc4[atc4[, paste0("significant_", atc4_table$var[i])],
                c("Uniprot", "Genes",
                  paste0("coef_", atc4_table$var[i]),
                  paste0("p.adj_", atc4_table$var[i]),
                  paste0("effect_size_", atc4_table$var[i]))]
    tab <- tab[order(tab[, paste0("p.adj_", atc4_table$var[i])]), ]
    colnames(tab) <- c("$UniProt$", "$Gene$", "$coef$", "$p_{adj}$",
                       "$ES$")
    rownames(tab) <- NULL
    pandoc.table(
        tab, split.table = "Inf", justify = "ccrrr", keep.line.breaks = FALSE,
        split.cells = 50, style = "rmarkdown", digits = 3,
        caption = paste0(
            "Proteins significantly associated with ATC level 4 medication *",
            atc4_table$Name[i], "* (ATC4 ", atc4_table$ATC[i], ").",
            " $coef$ and $p_{adj}$: coefficient (representing the ",
            "differential abundance in log2 scale) and p-value ",
            "adjusted for multiple hypothesis testing. $ES$ effect size. ",
            "Proteins are ordered by p-value."))
}
```

**Table S45** (external spreadsheet, xlsx format): Comparison with results from literature.


# Extended Material and Methods

\setcounter{table}{45}

```{=latex}
\begin{table}[h!]
\caption{Set up for gradient separation (left) and wash and equilibration (right).}
\begin{tabular}{llll|llll}
\multicolumn{4}{c|}{Gradient pump} & \multicolumn{4}{c}{Regeneration pump} \\ \hline
Time  & \%A & \%B & FR {[}ml/mn{]} & Time  & \%A  & \%B  & FR {[}ml/mn{]}  \\
0     & 97  & 3   & 0.8            & 0     & 97   & 3    & 0.8             \\
0.7   & 20  & 80  & 1              & 5     & 65   & 35   & 0.8             \\
1.5   & 20  & 80  & 1              & 5.05  & 97   & 3    & 1               \\
2     & 97  & 3   & 1              & 5.75  & 97   & 3    & 1               \\
5     & 97  & 3   & 1              & 5.8   & 97   & 3    & 0.8             \\
5.2   & 97  & 3   & 0.8            &       &      &      &
\end{tabular}
\end{table}
```
